Whole-chromosome pairwise alignment using the hierarchical wavefront algorithm

### 2021-06-14 manuscript/main.tex
a bit more explanation; citation

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
    to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and
    pangenome construction~\citep{33177663, 33066802}. Pairwise sequence alignment can be applied to identify
    similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome, chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. This opportunity, in combination with the massive amount of data
    available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.

    Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory
    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing methods
    like PacBio HiFi. To scale to vertebrate genomes, methods for pangenome construction based on these approaches
    must first filter out centromeric and other highly-repetitive sequences.

@@

    Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, incorporating
    the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
    Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
    runtime and memory-usage without affecting accuracy~\citep{32657365}.

    %TODO THINGS ON THE FILTERING?
    %TODO THING ON THE SPLIT READ AND/OR THE MERGING?

    In addition to the segment length, the minimum segment estimated identity and the maximum number of
    mappings to report for each segment can be specified in the mapping process. These settings allow users
    to precisely define the mapping space to consider, specifying the characteristics of homologies to compute.
    Each mapping location is then used as a target for base-level alignment using a hierarchical implementation
    of the WFA.

@@

    To avoid such limitation, \wfmash\ applies a hierarchical WFA, exploiting the WFA to guide the alignment
    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
    two fragments.

@@

    Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments
    on the optimal path in the high order DP-matrix. During the process, a further step is performed for
    refining the alignment. Indeed, inaccurate mapping estimation can lead to under-alignment at the beginning
    and the end of the aligned syntenic region. Therefore, edlib is applied by computing semi-global alignments
    to resolve the mapping boundaries.

    The hierarchical implementation requires only the memory to align a syntenic region at \textit{W}-bps resolution,
    limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
    This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
    no heuristic seeding step, and FIXME can benefit from parallel exploration of the wavefront.
    %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
    %TODO 'and can benefit from parallel exploration of the wavefront.' Can we?

@@

    \subsection{Simulated data}
    We simulated long sequences from the telomere-to-telomere (T2T) assemblies of the human chromosomes 8,
    at different levels of length and divergence (Table 1 and Table 2). Chromosomes were split into shorter
    sequences using the splitfa tool~\citep{splitfa}, In each run, all sequences have the same length, with 50\%
    of them in reverse complement. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs)
    were introduced using the Mutation-Simulator script~\citep{32780803}.
    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.

    To evaluate the correctness of the alignments, we used the pairwise alignment of the input sequences to build
    a pangenome graph in each run, identifying the variants embedded in the graphs. The graphs were built using~\citep{seqwish}.
    seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
    sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
    evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval}.

@@

    % TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000)

    % TODO Centromere (chr8:42,881,543-47,029,467)

    % TODO Simulation with full chromosomes

    \subsection{Real data}
    % HPRC data time/memory
    % Histogram of the contig length distribution (supplementary?)

>>>>

    Many biological analyses take advantage of aligning DNA sequences, ranging from read~mapping~\citep{22388286, BWA_MEM, 23103880}
    to variant detection~\citep{21478889}, as well as de~novo assembly~\citep{19251739} and
    pangenome construction~\citep{33177663, 33066802}. Pairwise sequence alignment can be applied to identify
    similar regions that may indicate functional, structural, and/or evolutionary
    relationships between two biological sequences. In this context, pangenomes model the full set
    of genomic elements in a given species or clade~\citep{32453966}. These data structures encode the mutual
    relationships between all the genomes represented, in contrast to reference-based approaches which
    relate sequences to a particular genome chosen as reference. The unbiased approach allows studying
    the entire genetic diversity of a population. This opportunity, in combination with the massive amount of data
    available nowadays, presses for the development of tools able to compute pairwise alignment of very long sequences.

    Popular seed-and-extend mapping methods, like Minimap2~\citep{29750242}, poorly scale in runtime and memory
    when generating sensitive alignments for chromosome-scale contigs available thanks to the new sequencing
    technologies~\citep{33288905}. To scale to vertebrate genomes, methods for pangenome construction based on these
    approaches must first filter out centromeric and other highly-repetitive sequences, in contrast to the pangenome
    idea of modeling the full genetic variation in the samples.

@@

    Each query sequence is broken into non-overlapping pieces of the requested length. These segments are then
    mapped using MashMap's sliding MinHash mapping algorithm~\citep{30423094}. We extended MashMap, incorporating
    the robust winnowing~\citep{Schleimer S. et al. (2003)} in the minimizer downsampling.
    Robust sampling avoids taking too many minimizers in low-complexity substrings, yielding improvements in
    runtime and memory-usage without affecting accuracy~\citep{32657365}. Futhermore, we made it possible
    to set the number of mappings to return for each segment; this is useful for identifying paralogous regions and
    homologous relationship between sets of genomes. Together with the segment length and the minimum segment
    estimated identity, these settings allow users to precisely define the mapping space to consider,
    specifying the characteristics of homologies to compute.
    \\

    FIXME THINGS ON THE FILTERING???\\

    FIXME THING ON THE SPLIT READ AND/OR THE MERGING???\\

    Finally, each mapping location is used as a target for base-level alignment using a hierarchical implementation
    of the WFA.

@@

    To avoid such limitation, \wfmash\ applies a hierarchical WFA, exploiting the WFA to guide the alignment
    process, but keeping the largest alignment problem size small. Rather than directly aligning the whole
    sequences, it aligns them to each other in small pieces \textit{W}-bps long by applying a global WFA alignment
    (Figure~1\vphantom{\ref{fig:1}}). The whole global alignment is then computed over the full
    dynamic-programming matrix (high order DP-matrix) at \textit{W}-bps resolution.
    Each cell in the high order DP-matrix corresponds to the alignment of a specific pair of fragments
    \textit{W}-bps long from the two sequences to align. To determine if each cell is a match, and then guiding the
    \textit{W}-bps resolution alignment, the global alignment using the standard WFA is performed between the
    two fragments. To accelerate the process, an alignment-free comparison of the two fragments is performed first by
    computing the mash distance between them. Cells whose fragments have a high mash distance are considered as mismatches,
    avoiding computing the corresponding base-level alignments.

@@

    Finally, the \textit{W}-bps resolution traceback is applied to determine the set of base-level alignments
    on the optimal path in the high order DP-matrix. During the process, a further step is performed for
    refining the alignment. Inaccurate mapping estimation can lead to under-alignment at the beginning
    and the end of the aligned syntenic region. Therefore, FIXME edlib is applied by computing semi-global alignments
    to resolve the mapping boundaries. Moreover, the standard WFA is applied to resolve breakpoints of the structural
    variants when present in the aligned region.

    The hierarchical implementation requires only the memory to align the sequences at \textit{W}-bps resolution,
    limiting the runtime and the memory of the standard WFA by applying it to fragments \textit{W}-bps long.
    This approach is more FIXME flexible than using a fixed-width band (it effectively has a variable bandwidth), requires
    no heuristic seeding step, and can benefit from parallel exploration of the wavefront.
    %TODO 'this approach is more flexible than using a fixed-width band: Not sure about writing it. I would remove it, else we should demonstrate it.
    %TODO 'and can benefit from parallel exploration of the wavefront.' Can we?

@@

    \subsection{Simulated data}
    We simulated long sequences from the \textit{Saccharomyces cerevisiae} chromosome IV (S288C strain) and the
    telomere-to-telomere (T2T) assemblies of the human chromosomes 8, at different levels of length and divergence
    (Table 1 and Table 2). Chromosomes were split into shorter sequences using the splitfa tool (6a0bb67)~\citep{splitfa}.
    In each run, all sequences have the same length, with 50\% of them in reverse complement respect to the source
    chromosome. Single nucleotide variants~(SNVs), small indels, and structural variants~(SVs) were introduced
    using the Mutation-Simulator script~\citep{32780803}.
    %, while structural variants (SVs) were applied using the SURVIVOR benchmarking tool~\citep{28117401}.

    To evaluate the correctness of the alignments, in each run we used the pairwise alignment of the input sequences to build
    a pangenome graph and identify the variants embedded in it. The graphs were built using~\citep{seqwish}.
    seqwish implements a lossless conversion from pairwise alignments between sequences to a pangenome graph encoding the
    sequences and their alignments. We called SNVs, small indels, and SVs using vg deconstruct~\citep{30125266},
    evaluating the false-negative and false-positive rates using vcfeval~\citep{vcfeval} for SNVs and small indels,
    and FIXME XXXXX for SVs.

@@

    % TODO Simulation with full chromosomes too

    % TODO 7 Mbp beta-defensin locus (chr8: 6,300,000-13,300,000): wfmash/minimap2/winnomnap2

    % TODO Centromere (chr8:42,881,543-47,029,467): only wfmash

    \subsection{Real data}
    % Yeast/HPRC data time/memory
    % Histogram of the contig length distribution (supplementary?)
